### 2025-11-09 sections/1-introduction.tex
brought back Partition

Given a distance function for the data, $k$-nearest neighbors ($k$-NN) search finds the $k$ points in a dataset that are closest to a given query point.
We use the term \textit{approximate} to refer to algorithms that have imperfect recall, while \textit{exact} refers to algorithms that have perfect recall as measured against a brute-force search.
Existing fast algorithms are often approximate~\cite{gao2023high}, which may suffice for some applications, but the need for fast \textit{and} exact search remains~\cite{ukey2023survey}.

This paper introduces CAKES (CLAM-Accelerated $K$-NN Entropy-Scaling Search), a set of three novel algorithms for exact $k$-NN search.
We benchmark these against FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We also evaluate performance on SILVA 18S~\cite{10.1093/nar/gks1219} using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and RadioML~\cite{oshea2018radioml} using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.
To further contextualize results, we also use a synthetic dataset with simple statistical properties.

@@

Entropy-scaling algorithms exploit the inherent manifold structure of large datasets, achieving complexity that scales with topological properties, such as metric entropy and fractal dimension, rather than cardinality.
This makes them especially suitable for manifold-constrained data.

In this paper, we introduce CAKES, a set of three entropy-scaling algorithms for $k$-NN search, implemented in Rust, extending CHESS~\cite{ishaq2019clustered} to $k$-NN search.
Sections~\ref{sec:methods:knn-search:repeated-rnn-complexity} and~\ref{sec:methods:knn-search:complexity-of-sieve-methods} provide a theoretical analysis.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a random uniform distribution.
Given that CAKES's algorithms are intended for datasets with a manifold structure, an analysis assuming a uniform distribution would be uninformative.
As a result, our analysis assumes datasets with manifold structure, reflecting the conditions for which CAKES is designed.

>>>>

Given a distance function for the data, $k$-nearest neighbors ($k$-NN) search finds the $k$ points in a dataset that are closest to a given query point.
We use the term \textit{approximate} to refer to algorithms that have imperfect recall, and \textit{exact} to refer to algorithms that have perfect recall as measured against a brute-force search.
Existing fast algorithms are often approximate~\cite{gao2023high}, which may suffice for some applications, but the need for \textit{fast and exact} search remains~\cite{ukey2023survey}.

This paper introduces CAKES (CLAM-Accelerated $K$-NN Entropy-Scaling Search), a set of three novel algorithms for exact $k$-NN search.
We benchmark CAKES against FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We also evaluate performance on SILVA 18S~\cite{10.1093/nar/gks1219} using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and RadioML~\cite{oshea2018radioml} using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.
To further contextualize results, we also use a synthetic dataset with simple statistical properties.

@@

Entropy-scaling algorithms exploit the inherent manifold structure of large datasets, achieving performance that scales with topological properties, such as metric entropy and fractal dimension, rather than cardinality.
This makes them especially suitable for manifold-constrained data.

CAKES is a set of three algorithms for exact $k$-NN search: Repeated $\rho$-NN, Breadth-First Sieve, and Depth-First Sieve.
In Section~\ref{sec:methods}, we describe the clustering and search algorithms, and provide a theoretical analysis.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a random uniform distribution.
Since CAKES is designed for datasets with a manifold structure, an analysis assuming a uniform distribution would be uninformative.
Therefore, our analysis assumes datasets with manifold structure, reflecting the conditions for which CAKES is designed.

### 2025-11-09 sections/2-methods.tex
brought back Partition

We present CAKES, a set of three novel algorithms for exact $k$-NN search in finite-dimensional spaces.
Given a \textit{dataset} $\textbf{X} = \{x_1, \dots, x_n\}$ and a \textit{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$, $k$-NN search aims to find the $k$ points in $\textbf{X}$ closest to a given \textit{query} point $q$ according to $f$.
If $f$ is a distance metric (i.e. it obeys the triangle inequality), then CAKES guarantees exact $k$-NN search.

CAKES assumes the manifold hypothesis~\cite{fefferman2016testing}, the notion that high($D$)-dimensional data collected from constrained generating phenomena often occupy a low($d$)-dimensional manifold within their embedding space ($d \ll D$).
We define the \emph{local fractal dimension} (LFD) to quantify this property.
For a point $q \in \textbf{X}$ and radii $r_1 > r_2$,
\begin{equation}
    \text{LFD}(q, r_1, r_2) = \frac{\text{log} \Big( \frac{|B(q, r_1)|}{|B(q, r_2)|} \Big) }{\text{log} \big( \frac{r_1}{r_2} \big) }
    \label{eq:methods:lfd-original}
\end{equation}
where $B(q, r)$ is the set of points within radius $r$ from $q$.

When the majority of points in the dataset have low ($d \ll D$) LFD, we simply say that the dataset has low LFD.
For real-world datasets, we expect the LFD to be locally uniform, i.e.,\,when $r$ is small, but (often) globally variable.

\subsection{Clustering}
\label{sec:methods:clustering}

We define a \emph{cluster} $C$ as a set of points with a \emph{center} and \emph{radius}.
The center $c$ is the geometric median of the points in the cluster (i.e., the point minimizing the sum of distances to all other points in $C$) and the radius $r$ is the maximum distance from $c$ to any point in $C$.
A cluster with only one point is a $leaf$, otherwise a $parent$ cluster has two $child$ clusters.
Starting from the root cluster $\mathcal{R}$ containing the entire dataset, Alg.~\ref{alg:methods:partition} recursively builds the tree $\mathcal{T}$ in $\mathcal{O}(n \log n)$ time.

\begin{figure}[t]
    \vskip -0.1in
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.pdf}
    \caption{
        {\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.
        ${\color{blue}\delta} = f(q, c)$ is the distance from the query to the cluster center $c$.
        ${\color{red}\delta^{+}} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
        ${\color{green}\delta^{-}} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
    }
    \label{fig:methods:deltas}
    \vskip -0.05in
\end{figure}

\begin{algorithm} % enter the algorithm environment
    \caption{\textsc{Partition}($C$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \Require $f$, a distance function
        \Require $C$, a cluster
        \Require $criteria$, user-specified continuation criteria

        \If{$|C| > 1$ \textbf{and} $C$ satisfies $criteria$}
        \State $S \Leftarrow$ random sample of $\left\lceil \sqrt{|C|} \right\rceil$ points from $C$
        \State $c \Leftarrow$ geometric median of $S$
        \State $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
        \State $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
        \State $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
        \State $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$
        \State \Call{Partition}{L, criteria}
        \State \Call{Partition}{R, criteria}
        \EndIf
    \end{algorithmic}
\end{algorithm}

@@

Given a query $q$ and integer $k$, $k$-NN search finds the $k$ closest points to $q$ in $\textbf{X}$. In this section, we present three novel algorithms for exact $k$-NN search:
Repeated $\rho$-NN, Breadth-First Sieve, and Depth-First Sieve.
We use $Q$ for the data structure containing the clusters that are still in contention for being among the $k$ nearest neighbors, and $H$ for a max-priority queue (of size $k$) containing the current $k$ nearest neighbors found.
These algorithms also use some terminology defined and illustrated in Figure~\ref{fig:methods:deltas}.

@@

We begin using \textsc{TreeSearch} on the root cluster $R$ with the search radius $r$ equal to the radius of $\mathcal{R}$ divided by $|X|$.
If the returned set $Q$ is empty, we double $r$ and repeat \textsc{TreeSearch} until $Q$ is non-empty.
Now, so long as $\sum_{C \in Q} |C| < k$, we repeat \textsc{TreeSearch}, but instead of doubling $r$, we multiply it by a factor determined by the LFD in the vicinity of the query ball.
In particular, we increase the radius by a factor of $\big( {\frac{k}{\sum_{C \in Q} |C|}} \big)^\frac{1}{\mu}$
where $\mu$ is the harmonic mean of the LFD of the clusters in $Q$ to ensure that this factor is not dominated by outlier clusters with very high LFD.
Intuitively, this factor will be large when we have found few points or the LFD is low because we need to increase the radius quickly, and small when we have found almost all $k$ points or LFD is large and we need to increase the radius slowly.
Once $\sum_{C \in Q} |C| \geq k$, we perform \textsc{LeafSearch} to find the $k$ nearest neighbors.

\begin{algorithm}[!t]
    \caption{\textsc{TreeSearch}($C$, $q$, $r$)}
    \label{alg:methods:rnn-search:tree-search}
    \begin{algorithmic}
        \Require $C$, a cluster
        \Require $q$, a query
        \Require $r$, a search radius
        \If{$\delta^+_C \leq r$}
            \State \textbf{return} $\{C\}$
        \EndIf
        \State $[L, R]$ $\Leftarrow$ \textit{children} of $C$
        \State \textbf{return} \Call{TreeSearch}{$L, q, r$} $\cup$ \Call{TreeSearch}{$R, q, r$}
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[!t]
    \caption{\textsc{LeafSearch}($Q$, $q$, $r$)}
    \label{alg:methods:rnn-search:leaf-search}
    \begin{algorithmic}
        \Require $f$, a distance function
        \Require $Q$, a set of clusters
        \Require $q$, a query
        \Require $r$, a search radius
        \State $H \Leftarrow \emptyset$
        \For{$C \in Q$}
            \If{$\delta^+_C \leq r$}
                \State $H$ $\Leftarrow$ $H \cup C$
            \Else
                \For{$p \in C$ \textbf{if} $f(p, q) \leq r$}
                    \State $H$ $\Leftarrow$ $H \cup \{p\}$
                \EndFor
            \EndIf
        \EndFor
        \State \textbf{return} $H$
    \end{algorithmic}
\end{algorithm}

@@

For \textsc{LeafSearch}, we must estimate $\sum_{C \in Q} |C|$;
since we must examine every point in each such cluster, time complexity of \textsc{LeafSearch} is linear in this quantity.
Let $\rho_k$ be the distance from the query to the $k^{th}$ nearest neighbor.
Then, we see that $Q$ is the set of clusters that overlap with a ball of radius $\rho_k$ around the query.
We can estimate this region as a ball of radius $\rho_k + 2\hat{r}$, where $\hat{r}$ is the mean radius of the clusters in $Q$.
Thus, the \textsc{LeafSearch} component of Equation~\ref{eq:methods:rnn-search-complexity} becomes $k \cdot \left( \frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} \right)^{\hat{d}}$, where $\hat{d}$ is the harmonic mean of LFDs in $Q$.
It remains to estimate $\rho_k$.

While ordinarily we would compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|C|}$, of clusters in $Q$ at a radius equal to the mean of their radii, $\hat{r}$.
Since $q$ is from the same distribution as $X$, the LFD at $q$ should not be significantly different from that at the center of a cluster in $Q$.
By Equation~\ref{eq:methods:lfd-original}, we have $\hat{d} = \frac{\log{}\frac{\hat{|C|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$, and rearranging, $\frac{\hat{r}}{\rho_k} = \left( \frac{\hat{|C|}}{k} \right)^{\hat{d}^{-1}}$.
This simplifies the \textsc{LeafSearch} component to $k \Bigg( 1 + 2 \cdot \bigg( \frac{\hat{|C|}}{k} \bigg) ^ {\hat{d}^{-1}} \Bigg)^d$.

Again using the assumption that $q$ is from the same distribution as $X$, we have that $\hat{d} \approx d$.
Thus, the asymptotic complexity of Repeated $\rho$-NN is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            k \cdot
            \overbrace{\bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg).
    \label{eq:methods:repeated-rnn-complexity}
\end{gather}

@@

We begin by letting $Q$ be a list of 3-tuples $(C, \delta^{+}, m)$, where $C$ is a cluster, $\delta^{+}$ is as illustrated in Figure~\ref{fig:methods:deltas}, and $m$ is the multiplicity of $C$ in $Q$.
$Q$ starts containing only $(\mathcal{R}, \delta^{+}_{\mathcal{R}}, |\mathcal{R}| - 1)$ while $H$ starts with the center of $\mathcal{R}$.

We then use the QuickSelect algorithm, modified to account for multiplicities and to reorder $Q$ in-place, to find $\tau$, the smallest $\delta^{+}$ in $Q$ such that $\left| B(q, \tau) \right| \geq k$ (including the points in $H$).
Since this step may require a binary search of $Q$ and reordering $Q$ takes linear time, this version of QuickSelect has $\mathcal{O}\left(|Q| \log |Q|\right)$ time complexity.

@@

This algorithm (DFS) greedily descends into the child cluster whose center if closer to the query than that of its sibling.

Let $Q$ be a min-queue of clusters prioritized by $\delta^{-}$ starting with only $\mathcal{R}$ while $H$ starts empty.
So long as $H$ is not full or the top-priority items from $H$ and $Q$ have $\delta_H \geq \delta_Q$, we repeat following:
\begin{itemize}
    \item While the top-priority cluster is not a leaf, remove it from $Q$ and add its children to $Q$.
    \item Remove the top-priority cluster (a leaf) from $Q$ and add all its points to $H$.
\end{itemize}
This process terminates when $H$ is full and the top-priority items in $H$ and $Q$ are such that $\delta_H < \delta_Q$, i.e.,\,the next closest point to be visited is farther from the query than the $k^{th}$ nearest neighbor found so far.
Thus, $H$ contains the $k$ nearest neighbors to the query.

@@

The asymptotic complexity of BFS is dominated by our modified QuickSelect algorithm, which accounts for the clusters in $Q$ and the points in $H$, and is log-linear in their sum.
Thus, the asymptotic complexity of BFS is
\begin{equation}
    \mathcal{O} \big( (T + L) \log(T + L) \big).
    \label{eq:methods:bfs-complexity}
\end{equation}
Similarly, the asymptotic complexity of DFS is dominated by the priority queue operations for $T$ clusters and $L$ points;
it is given by
\begin{equation}
    \mathcal{O} \big( T \log T + L \log L \big).
    \label{eq:methods:dfs-complexity}
\end{equation}

>>>>

Given a dataset $\textbf{X} = \{x_1, \dots, x_n\}$ and a distance function $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$, we define a \emph{cluster} $C$ as a set of points with a \emph{center} and \emph{radius}.
The center $c$ is the geometric median of the points in the cluster (i.e., the point minimizing the sum of distances to all other points in $C$) and the radius $r$ is the maximum distance from $c$ to any point in $C$.
Using previous work in CHESS~\cite{ishaq2019clustered}, and CLAM~\cite{ishaq2021clustered}, we build a binary tree $\mathcal{T}$ of clusters by recursively partitioning the dataset $\textbf{X}$ (See Alg.~\ref{alg:methods:partition}).
In practice, we approximate $c$ using a random sample of points from $C$.
\begin{algorithm}[h] % enter the algorithm environment
    \caption{\textsc{Partition}($C$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \Require $f$, a distance function
        \Require $C$, a cluster
        \Require $criteria$, user-specified continuation criteria

        \If{$|C| > 1$ \textbf{and} $C$ satisfies $criteria$}
        \State $c \Leftarrow$ geometric median of $C$
        \State $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
        \State $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
        \State $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
        \State $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$
        \State \Call{Partition}{L, criteria}
        \State \Call{Partition}{R, criteria}
        \EndIf
    \end{algorithmic}
\end{algorithm}

The manifold hypothesis~\cite{fefferman2016testing} states that high($D$)-dimensional data collected from constrained generating phenomena occupy a low($d$)-dimensional manifold within their embedding space ($d \ll D$).
We define the \emph{local fractal dimension} (LFD) to quantify this property:
\begin{equation}
    \text{LFD}(p, r_1, r_2) = \frac{\text{log} \big( \frac{|B(p, r_1)|}{|B(p, r_2)|} \big) }{\text{log} \big( \frac{r_1}{r_2} \big) }
    \label{eq:methods:lfd-original}
\end{equation}
where $p \in X$ is a point and $B(p, r)$ is the set of points within radius $r$ from $p$.
When $d \ll D$ for the majority of points, we simply say that the dataset has low LFD.

@@

$k$-NN search in CAKES guarantees perfect recall when $f$ is a metric (i.e.\ it satisfies the triangle inequality) with performance that scales with LFD rather than $|X|$.

In the following, we use $Q$ for the data structure containing clusters that are in contention for containing the $k$ nearest neighbors and $H$ for a max-priority queue (of size $k$) containing the current $k$ nearest neighbors found.
We denote the distances from the query to the center, the theoretically farthest, and the theoretically closest points in a cluster $C$ using $\delta_C$, $\delta^{+}_C$, and $\delta^{-}_C$, respectively.
We also assume that the query $q$ is drawn from the same distribution as $X$.

@@

We begin by using \textsc{TreeSearch} on the root cluster $R$ with the search radius $r$ equal to the radius of $\mathcal{R}$ divided by $|X|$.
If the returned set $Q$ is empty, we double $r$ and repeat \textsc{TreeSearch} until $Q$ is non-empty.
Now, so long as $\sum_{C \in Q} |C| < k$, we repeat \textsc{TreeSearch}, but instead of doubling $r$, we multiply it by a factor determined by the LFD in the vicinity of the query ball.
In particular, we increase the radius by a factor of $\big( {\frac{k}{\sum_{C \in Q} |C|}} \big)^\frac{1}{\mu}$
where $\mu$ is the harmonic mean of the LFD of the clusters in $Q$ to ensure that this factor is not dominated by outlier clusters with very high LFD.
Intuitively, this factor will be large when we have found few points or the LFD is low because we need to increase the radius quickly, and small when we have found almost all $k$ points or LFD is large and we need to increase the radius slowly.
Once $\sum_{C \in Q} |C| \geq k$, we perform \textsc{LeafSearch} to find the $k$ nearest neighbors.

\begin{algorithm}[h]
    \caption{\textsc{TreeSearch}($C$, $q$, $r$)}
    \label{alg:methods:rnn-search:tree-search}
    \begin{algorithmic}
        \Require $C$, a cluster
        \Require $q$, a query
        \Require $r$, a search radius
        \If{$\delta^+_C \leq r$}
            \State \textbf{return} $\{C\}$
        \EndIf
        \State $[L, R]$ $\Leftarrow$ \textit{children} of $C$
        \State \textbf{return} \Call{TreeSearch}{$L, q, r$} $\cup$ \Call{TreeSearch}{$R, q, r$}
    \end{algorithmic}
\end{algorithm}
\vspace{-0.2in}

\begin{algorithm}[h]
    \caption{\textsc{LeafSearch}($Q$, $q$, $r$)}
    \label{alg:methods:rnn-search:leaf-search}
    \begin{algorithmic}
        \Require $f$, a distance function
        \Require $Q$, a set of clusters
        \Require $q$, a query
        \Require $r$, a search radius
        \State $H \Leftarrow \emptyset$
        \For{$C \in Q$}
            \For{$p \in C$ \textbf{if} $f(p, q) \leq r$}
                \State $H$ $\Leftarrow$ $H \cup \{p\}$
            \EndFor
        \EndFor
        \State \textbf{return} $H$
    \end{algorithmic}
\end{algorithm}
\vspace{-0.2in}

@@

For \textsc{LeafSearch}, we must estimate $\sum_{C \in Q} |C|$.
Since we must examine every point in each such cluster, the time complexity of \textsc{LeafSearch} is linear in this quantity.
Let $\rho_k$ be the distance from the query to the $k^{th}$ nearest neighbor.
Then, we see that $Q$ is the set of clusters that overlap with a ball of radius $\rho_k$ around the query.
We can estimate this region as a ball of radius $\rho_k + 2\hat{r}$, where $\hat{r}$ is the mean radius of the clusters in $Q$.
Thus, the \textsc{LeafSearch} component of Equation~\ref{eq:methods:rnn-search-complexity} becomes $k \cdot ( \frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} )^{\hat{d}}$, where $\hat{d}$ is the harmonic mean of LFDs in $Q$.
It remains to estimate $\rho_k$.

While ordinarily we would compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|C|}$, of clusters in $Q$ at a radius equal to the mean of their radii, $\hat{r}$.
Assuming that $q$ is from the same distribution as $X$, by Equation~\ref{eq:methods:lfd-original}, we have $\hat{d} = \frac{\log{}\frac{\hat{|C|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$, and rearranging, $\frac{\hat{r}}{\rho_k} = ( \frac{\hat{|C|}}{k} )^{\hat{d}^{-1}}$.
This simplifies the \textsc{LeafSearch} component to $k ( 1 + 2 \cdot ( \frac{\hat{|C|}}{k} ) ^ {\hat{d}^{-1}} )^d$.

Again using the assumption that $q$ is from the same distribution as $X$, we have that $\hat{d} \approx d$.
Thus, the asymptotic complexity of Repeated $\rho$-NN is
\begin{gather}
    \mathcal{O}
    \bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            k \cdot
            \overbrace{\Big( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \Big)^d}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \bigg).
    \label{eq:methods:repeated-rnn-complexity}
\end{gather}

@@

We begin by letting $Q$ be a list of 3-tuples $(C, \delta^{+}, m)$, where $m$ is the multiplicity of $C$ in $Q$.
$Q$ starts containing only $(\mathcal{R}, \delta^{+}_{\mathcal{R}}, |\mathcal{R}| - 1)$ while $H$ starts with the center of $\mathcal{R}$, where $\mathcal{R}$ is the root of $\mathcal{T}$.
We then use QuickSelect, modified to account for multiplicities and to reorder $Q$ in-place, to find $\tau$, the smallest $\delta^{+}$ in $Q$ such that $\left| B(q, \tau) \right| \geq k$ (including the points in $H$).
Since this step may require a binary search of $Q$ and reordering $Q$ takes linear time, this version of QuickSelect has $\mathcal{O}\left(|Q| \log |Q|\right)$ time complexity.

@@

Let $Q$ be a min-queue of clusters prioritized by $\delta^{-}$ starting with only $\mathcal{R}$ while $H$ starts empty.
So long as $H$ is not full or the top-priority items from $H$ and $Q$ have $\delta_H \geq \delta_Q$, we repeat following:
\begin{itemize}
    \item While the top-priority cluster is not a leaf, remove it from $Q$ and add its children to $Q$.
    \item Remove the top-priority cluster (a leaf) from $Q$, add all its points to $H$, and trim $H$ to size $k$.
\end{itemize}
This process terminates when $|H| = k$ and the top-priority items in $H$ and $Q$ are such that the next closest point to be visited in $Q$ is farther from the query than the $k^{th}$ nearest neighbor found so far.
Thus, $H$ contains the $k$ nearest neighbors to the query.

@@

The asymptotic complexity of BFS is dominated by our modified QuickSelect algorithm, which accounts for the clusters in $Q$ and the points in $H$, and is log-linear in their sum.
Similarly, that of DFS is dominated by the priority queue operations for $T$ clusters and $L$ points.
These complexities are:
\begin{equation*}
    % \label{eq:methods:sieve-complexity}
    \begin{split}
    BFS \sim \ & \mathcal{O} \big( (T + L) \log(T + L) \big) \\
    DFS \sim \ & \mathcal{O} \big( T \log T + L \log L \big)
    \end{split}
\end{equation*}

### 2025-11-09 sections/3-datasets.tex
down to six pages
brought back Partition

\subsection{Datasets}
\label{sec:datasets-and-benchmarks:ann-benchmark-datasets}

@@

We also include benchmarks on the SilVA-18S and Radio-ML datasets.
For the SilVA-18S dataset, which consists of 2,224,640 sequences of up-to 3,712 nucleotides, we use the Levenshtein distance metric~\cite{levenshtein1966binary}.
The Radio-ML dataset consists of 1,024-dimensional complex-valued time-series samples, spanning 24 modulation modes and 26 SNR levels from $-20$dB to $30$dB.
We use the 10dB SNR subset with 97,304 samples and Dynamic Time Warping~\cite{muller2007dynamic} as the distance metric.
For both datasets, we hold out 1,000 random samples as queries and build the tree on the remaining samples.
These were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM running Manjaro Linux 5.15.164-1-MANJARO.

\begin{table}
    \caption{Datasets used in benchmarks.}
    \label{tab:datasets:summary}
    \begin{center}
        \begin{sc}
            \begin{tabular}{|l|l|l|l|}
                \hline
                \textbf{Dataset} & \textbf{Dist. Function}  &\textbf{Card}  & \textbf{Dim}  \\
                \hline
                Fashion-Mnist    & Euclidean                   & 60,000             & 784                    \\
                \hline
                Glove-25         & Cosine                      & 1,183,514          & 25                     \\
                \hline
                Sift             & Euclidean                   & 1,000,000          & 128                    \\
                \hline
                Random           & Euclidean                   & 1,000,000          & 128                    \\
                \hline
                SILVA            & Levenshtein                 & 2,224,640          & 3,712         \\
                \hline
                RadioML          & Dynamic Time Warping        & 97,920             & 1,024                  \\
                \hline
            \end{tabular}
        \end{sc}
    \end{center}
    \vskip -0.1in
\end{table}

@@

In order to study scaling behavior as cardinality increases, we synthetically augment the ANN-Benchmarks datasets by generating new points by applying small random perturbations to existing points in the original dataset.
Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, \ref{tab:results:qps-and-recall-random} show the results as the cardinality increases, with ``Mult.'' indicating the multiplier by which the cardinality was augmented.
Since the SilVA-18S and Radio-ML datasets are already large and are not vector datasets, we do not augment them;
instead, we subsample them to lower cardinalities for the scaling experiments.

@@

Since the time complexity of CAKES algorithms scales with the LFD of the dataset, we examine, in Figure~\ref{fig:results:lfd-plots}, the LFD of each dataset we used for benchmarks.

\begin{figure}[t]
    \centering
    % Row 1
    \subfloat[Fashion-MNIST\label{fig:results:fashion-mnist-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/fashion-mnist.png}}
    \hfill
    \subfloat[Glove-25\label{fig:results:glove-25-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/glove-25.png}}

    % Row 2
    \par\medskip
    \subfloat[Sift\label{fig:results:sift-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/sift.png}}
    \hfill
    \subfloat[Random dataset\label{fig:results:random-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/random.png}}

@@

Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, and \ref{tab:results:qps-and-recall-random} show the throughput (QPS) and recall for each algorithm on each of the vector datasets.
Since HNSW, ANNOY, and FAISS do not support the Levenshtein or DTW distance functions, we only benchmark CAKES on the SilVA-18S and RadioML datasets and report results in Figures~\ref{fig:results:silva-scaling} and~\ref{fig:results:radioml-scaling}.
All benchmarks were run with $k=10$, throughput is averaged over all queries, and recall is computed using a na\"ive linear scan.

>>>>

% \subsection{Datasets}
% \label{sec:datasets-and-benchmarks:ann-benchmark-datasets}

@@

We also include benchmarks on the SilVA-18S and Radio-ML datasets.
For the SilVA-18S dataset, which consists of 2,224,640 sequences of up-to 3,712 nucleotides, we use the Levenshtein distance metric~\cite{levenshtein1966binary}.
The Radio-ML dataset consists of 1,024-dimensional complex-valued time-series samples, spanning 24 modulation modes and 26 SNR levels from $-20$dB to $30$dB.
We use the 10dB SNR subset with 97,304 samples and Dynamic Time Warping~\cite{muller2007dynamic} as the distance metric.
For both datasets, we hold out 1,000 random samples as queries and build the tree on the remaining samples.
These benchmarks were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM running Manjaro Linux 5.15.164-1.

\begin{table}[h]
  \vskip -0.1in
  \caption{Datasets used in benchmarks.}
  \vskip -0.25in
  \label{tab:datasets:summary}
  \begin{center}
    \begin{sc}
      \begin{tabular}{|l|l|l|l|}
        \hline
        \textbf{Dataset} & \textbf{Dist. Function}  &\textbf{Card}  & \textbf{Dim}  \\
        \hline
        Fashion-Mnist    & Euclidean                   & 60,000             & 784                    \\
        \hline
        Glove-25         & Cosine                      & 1,183,514          & 25                     \\
        \hline
        Sift             & Euclidean                   & 1,000,000          & 128                    \\
        \hline
        Random           & Euclidean                   & 1,000,000          & 128                    \\
        \hline
        SILVA            & Levenshtein                 & 2,224,640          & 3,712         \\
        \hline
        RadioML          & Dynamic Time Warping        & 97,920             & 1,024                  \\
        \hline
      \end{tabular}
    \end{sc}
  \end{center}
  \vskip -0.25in
\end{table}

@@

In order to study how search performance scales as cardinality increases, we synthetically augment the ANN-Benchmarks datasets by generating new points by applying small random perturbations to existing points in the original dataset.
Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, \ref{tab:results:qps-and-recall-random} show the results as the cardinality increases, with ``Mult.'' indicating the multiplier by which the cardinality was augmented.
Since the SilVA-18S and Radio-ML datasets are already large and are not vector datasets, we do not augment them;
instead, we subsample them to lower cardinalities for the scaling experiments.

@@

Since the time complexity of CAKES algorithms scales with the LFD of the dataset, we examine, in Figure~\ref{fig:results:lfd-plots}, the LFD of each dataset we used for benchmarks.
\begin{figure}[h]
    \vskip -0.2in
    \centering
    % Row 1
    \subfloat[Fashion-MNIST\label{fig:results:fashion-mnist-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/fashion-mnist.png}}
    \hfill
    \subfloat[Glove-25\label{fig:results:glove-25-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/glove-25.png}}
    \vskip -0.2in

    % Row 2
    \par\medskip
    \subfloat[Sift\label{fig:results:sift-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/sift.png}}
    \hfill
    \subfloat[Random dataset\label{fig:results:random-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/random.png}}
    \vskip -0.2in

@@

Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, and \ref{tab:results:qps-and-recall-random} show the throughput (QPS) and recall for each algorithm on each of the vector datasets.
Since HNSW, ANNOY, and FAISS do not support the Levenshtein or DTW distance functions, we only benchmark CAKES on the SilVA-18S and RadioML datasets and report results in Figures~\ref{fig:results:silva-scaling} and~\ref{fig:results:radioml-scaling}.
All benchmarks were run with $k=10$, throughput is averaged over all queries, and recall is computed using a na\"ive linear scan as ground truth.
